Are populations like a circuit? Comparing isolation by resistance to a new coalescent-based method.

### 2019-04-20 isolation_by_coalescence.tex
updated figure scaling
erik comments

Equation \eqref{eqn:C_matrix} is linear in $G$ and $\gamma$,
and so after rearrangement can be solved with standard linear algebra
(and is related to the Sylvester equation \citep{bhatia1997solve}).
However, the solution is again not necessarily unique:
suppose that we have a matrix $Z$ such that $ZC$ is skew-symmetric.
Then $(G + Z) C + C (G + Z)^T = GC + CG^T$,
and so a Markov chain with generator matrix $G + Z$ has the same coalescence times
as the original chain.
Furthermore, for $G + Z$ to be a valid generator of a Markov chain,
the rows of $Z$ must sum to zero,
and all offdiagonal entries of $G + Z$ must be nonnegative. \revpoint{1}{11}
As before, if all entries of $G$ are nonzero, 
it is always possible to find sufficiently small $\epsilon$
so that $G + \epsilon Z$ remains the generator of a valid Markov chain.

@@

\paragraph{Other barriers: population sinks and recent expansions}
We simulated two other situations on a continuous landscape that produced an bias in effective migration.
First, we simulated a \textbf{valley} in a landscape with a $5 \times 3$ aspect ratio as above, 
in which fecundity of individuals in the middle 1/4 of the landscape was reduced by 50\%.
The valley is a population ``sink'',
although it would not be detectable with mortality rates,
and shows only a slightly lower population density (Supplemental Figure \ref{sfig:valley_results}).
Since individuals in the valley are more likely to have parents outside of the valley,
this produces gene flow ``uphill'', away from the ridge.
Next, we simulated a recent \textbf{expansion},
in which the middle 4/5 of the landscape was uninhabitable for 400,000 generations, \revpoint{1}{18}
until 500 generations before the present day.
(This is a model of secondary contact with a long period of isolation,
but not outside the realm of possibility.)
The resulting expansion produces gene flow back away from the center,
similar to the ``valley''.

@@

\begin{figure}
\centering
     \includegraphics[scale=0.8]{figs/5x3boxplots}
    \caption{Posterior distributions for values of $g$ 
    for the 5 $\times$ 3 graph with barriers 
    for each of the analysis cases with coalescence time inference.
    See Figure \ref{fig:5x3b_grid} to see movement parameter locations.
}
    \label{fig:5x3boxplots_mult_gam}
\end{figure}

\begin{figure}
\centering
     \includegraphics[scale=.7]{figs/4x4_grids}
    \caption{
        Grid structure and gene flow rates
        for the uniform graph, symmetric graph, the asymmetric graph,
        and the biased asymmetric graph 
        using the igraph R package.
    }
    \label{fig:4x4_grids}
\end{figure}

@@

\begin{figure}
\centering
    \includegraphics[width=0.45\textwidth]{figs/bias_coal_graph}
    \includegraphics[width=0.45\textwidth]{figs/bias_commute_graph}
    \includegraphics[width=0.45\textwidth]{figs/bias_coal_dists}
    \includegraphics[width=0.45\textwidth]{figs/bias_commute_dists}
\caption{
    \textbf{(Above:)} Posterior medians and \textbf{(below)} posterior distributions
    of the inferred values of $g$ 
    for the $4 \times 4$ discretization 
    of the square landscape with biased migration,
    inferred with the
    \textbf{(left)} coalescence time-based,
    and \textbf{(right)} resistance distance-based method. 
    In the boxplots,
    black boxes show posterior distributions of gene flow rates up and to the right,
    and the adjacent green box corresponds to gene flow rate along the same edge
    in the opposite direction. 
    }
\label{fig:grid_bias_4x4_1}
\end{figure}

\begin{figure}
\centering
    \includegraphics[width=0.45\textwidth]{figs/barrier_locations}
    \includegraphics[width=0.45\textwidth]{figs/barrier_eems}
    \includegraphics[width=0.45\textwidth]{figs/barrier_coal_graph}
    \includegraphics[width=0.45\textwidth]{figs/barrier_commute_graph}
    \includegraphics[width=0.45\textwidth]{figs/barrier_coal_dists}
    \includegraphics[width=0.45\textwidth]{figs/barrier_commute_dists}
    \caption{
        Results from the \textbf{barrier} simulation:
        \textbf{(top left)} Positions of all individuals at the end of simulation.
        \textbf{(top right)} mean migration rates as estimated by EEMS.
        Below are shown
        posterior median migration rates 
            as estimated using
            \textbf{(middle left)} the coalescence time method and
            \textbf{(middle right)} the resistance distance method.
        Last are shown the posterior distributions, for
            \textbf{(bottom left)} the coalescence time method and
            \textbf{(bottom right)} the resistance distance method.
        \label{sfig:barrier_results}
    }
\end{figure}

\begin{figure}
\centering
    \includegraphics[width=0.45\textwidth]{figs/valley_locations}
    \includegraphics[width=0.45\textwidth]{figs/valley_eems}
    \includegraphics[width=0.45\textwidth]{figs/valley_coal_graph}
    \includegraphics[width=0.45\textwidth]{figs/valley_commute_graph}
    \includegraphics[width=0.45\textwidth]{figs/valley_coal_dists}
    \includegraphics[width=0.45\textwidth]{figs/valley_commute_dists}
    \caption{
        Results from the \textbf{valley} simulation:
        \textbf{(top left)} Positions of all individuals at the end of simulation.
        \textbf{(top right)} mean migration rates as estimated by EEMS.
        Below are shown
        posterior median migration rates 
            as estimated using 
            \textbf{(middle left)} the coalescence time method and
            \textbf{(middle right)} the resistance distance method.
        Last are shown the posterior distributions, for
            \textbf{(bottom left)} the coalescence time method and
            \textbf{(bottom right)} the resistance distance method.
        \label{sfig:valley_results}
    }
\end{figure}

\begin{figure}
\centering
    \includegraphics[width=0.45\textwidth]{figs/expansion_locations}
    \includegraphics[width=0.45\textwidth]{figs/expansion_eems}
    \includegraphics[width=0.45\textwidth]{figs/expansion_coal_graph}
    \includegraphics[width=0.45\textwidth]{figs/expansion_commute_graph}
    \includegraphics[width=0.45\textwidth]{figs/expansion_coal_dists}
    \includegraphics[width=0.45\textwidth]{figs/expansion_commute_dists}
    \caption{
        Results from the \textbf{expansion} simulation:
        \textbf{(top left)} Positions of all individuals at the end of simulation.
        \textbf{(top right)} mean migration rates as estimated by EEMS.
        Below are shown
        posterior median migration rates 
            as estimated using 
            \textbf{(middle left)} the coalescence time method and
            \textbf{(middle right)} the resistance distance method.
        Last are shown the posterior distributions, for
            \textbf{(bottom left)} the coalescence time method and
            \textbf{(bottom right)} the resistance distance method.
        \label{sfig:expansion_results}
    }
\end{figure}

\begin{figure}
\centering
    \includegraphics[width=0.7\textwidth]{poplars/posterior_dists_g_populus}
    \includegraphics[width=0.7\textwidth]{poplars/grid_populus}
    \caption{
        Posterior distributions of parameters for \textit{Populus} data,
        using coalescence time inference with the model shown in Figure~\ref{fig:poplar_results}.
        \textbf{(top)} Posterior medians (dark line),
        interquartile range (box), and middle 95\% (whiskers),
        for gene flow rates.
        \textbf{(bottom)} Posterior medians, along with edge labels that match the upper figure.
        \label{sfig:poplar_g}
    }
\end{figure}

>>>>

Equation \eqref{eqn:C_matrix} is linear in $G$ and $\gamma$,
and so after rearrangement can be solved with standard linear algebra
(and is related to the Sylvester equation \citep{bhatia1997solve}).
However, the solution is again not necessarily unique:
suppose that we have a matrix $Z$ such that $ZC$ is skew-symmetric.
Then $(G + Z) C + C (G + Z)^T = GC + CG^T$,
and so a Markov chain with generator matrix $G + Z$ has the same coalescence times
as the original chain.
Furthermore, for $G + Z$ to be a valid generator of a Markov chain,
the rows of $Z$ must sum to zero,
and all offdiagonal entries of $G + Z$ must be nonnegative. \revpoint{1}{11}
As before, if all entries of $G$ are nonzero, 
it is always possible to find sufficiently small $\epsilon$
so that $G + \epsilon Z$ remains the generator of a valid Markov chain.
The coalescence rates, $\gamma$, provide additional degrees of freedom
which may render solutions nonunique in a broader range of situations.

@@

\paragraph{Other barriers: population sinks and recent expansions}
We simulated two other situations on a continuous landscape that produced an bias in effective migration.
First, we simulated a \textbf{valley} in a square landscape,
in which fecundity of individuals in the middle 1/4 of the landscape was reduced by 50\%.
The valley is a population ``sink'',
although it would not be detectable with mortality rates,
and shows only a slightly lower population density (Supplemental Figure \ref{sfig:valley_results}).
Since individuals in the valley are more likely to have parents outside of the valley,
this produces gene flow ``uphill'', away from the ridge.
Next, we simulated a recent \textbf{expansion},
in which the middle 4/5 of the landscape was uninhabitable for 400,000 generations, \revpoint{1}{18}
until 500 generations before the present day.
(This is a model of secondary contact with a long period of isolation,
but not outside the realm of possibility.)
The resulting expansion produces gene flow back away from the center,
similar to the ``valley''.

@@

\begin{figure}
\centering
     \includegraphics[scale=1]{figs/5x3boxplots}
    \caption{Posterior distributions for values of $g$ 
    for the 5 $\times$ 3 graph with barriers 
    for each of the analysis cases with coalescence time inference.
    See Figure \ref{fig:5x3b_grid} to see movement parameter locations.
}
    \label{fig:5x3boxplots_mult_gam}
\end{figure}

\begin{figure}
\centering
     \includegraphics[scale=1]{figs/4x4_grids}
    \caption{
        Grid structure and gene flow rates
        for the uniform graph, symmetric graph, the asymmetric graph,
        and the biased asymmetric graph 
        using the igraph R package.
    }
    \label{fig:4x4_grids}
\end{figure}

@@

\begin{figure}
\centering
    \includegraphics[width=0.49\textwidth]{figs/bias_coal_graph}
    \includegraphics[width=0.49\textwidth]{figs/bias_commute_graph}
    \includegraphics[width=0.49\textwidth]{figs/bias_coal_dists}
    \includegraphics[width=0.49\textwidth]{figs/bias_commute_dists}
\caption{
    \textbf{(Above:)} Posterior medians and \textbf{(below)} posterior distributions
    of the inferred values of $g$ 
    for the $4 \times 4$ discretization 
    of the square landscape with biased migration,
    inferred with the
    \textbf{(left)} coalescence time-based,
    and \textbf{(right)} resistance distance-based method. 
    In the boxplots,
    black boxes show posterior distributions of gene flow rates up and to the right,
    and the adjacent green box corresponds to gene flow rate along the same edge
    in the opposite direction. 
    }
\label{fig:grid_bias_4x4_1}
\end{figure}

\begin{figure}
\centering
    \includegraphics[width=0.4\textwidth]{figs/barrier_locations}
    \includegraphics[width=0.4\textwidth]{figs/barrier_eems}
    \includegraphics[width=0.4\textwidth]{figs/barrier_coal_graph}
    \includegraphics[width=0.4\textwidth]{figs/barrier_commute_graph}
    \includegraphics[width=0.4\textwidth]{figs/barrier_coal_dists}
    \includegraphics[width=0.4\textwidth]{figs/barrier_commute_dists}
    \caption{
        Results from the \textbf{barrier} simulation:
        \textbf{(top left)} Positions of all individuals at the end of simulation.
        \textbf{(top right)} mean migration rates as estimated by EEMS.
        Below are shown
        posterior median migration rates 
            as estimated using
            \textbf{(middle left)} the coalescence time method and
            \textbf{(middle right)} the resistance distance method.
        Last are shown the posterior distributions, for
            \textbf{(bottom left)} the coalescence time method and
            \textbf{(bottom right)} the resistance distance method.
        \label{sfig:barrier_results}
    }
\end{figure}

\begin{figure}
\centering
    \includegraphics[width=0.4\textwidth]{figs/valley_locations}
    \includegraphics[width=0.4\textwidth]{figs/valley_eems}
    \includegraphics[width=0.4\textwidth]{figs/valley_coal_graph}
    \includegraphics[width=0.4\textwidth]{figs/valley_commute_graph}
    \includegraphics[width=0.4\textwidth]{figs/valley_coal_dists}
    \includegraphics[width=0.4\textwidth]{figs/valley_commute_dists}
    \caption{
        Results from the \textbf{valley} simulation:
        \textbf{(top left)} Positions of all individuals at the end of simulation.
        \textbf{(top right)} mean migration rates as estimated by EEMS.
        Below are shown
        posterior median migration rates 
            as estimated using 
            \textbf{(middle left)} the coalescence time method and
            \textbf{(middle right)} the resistance distance method.
        Last are shown the posterior distributions, for
            \textbf{(bottom left)} the coalescence time method and
            \textbf{(bottom right)} the resistance distance method.
        \label{sfig:valley_results}
    }
\end{figure}

\begin{figure}
\centering
    \includegraphics[width=0.4\textwidth]{figs/expansion_locations}
    \includegraphics[width=0.4\textwidth]{figs/expansion_eems}
    \includegraphics[width=0.4\textwidth]{figs/expansion_coal_graph}
    \includegraphics[width=0.4\textwidth]{figs/expansion_commute_graph}
    \includegraphics[width=0.4\textwidth]{figs/expansion_coal_dists}
    \includegraphics[width=0.4\textwidth]{figs/expansion_commute_dists}
    \caption{
        Results from the \textbf{expansion} simulation:
        \textbf{(top left)} Positions of all individuals at the end of simulation.
        \textbf{(top right)} mean migration rates as estimated by EEMS.
        Below are shown
        posterior median migration rates 
            as estimated using 
            \textbf{(middle left)} the coalescence time method and
            \textbf{(middle right)} the resistance distance method.
        Last are shown the posterior distributions, for
            \textbf{(bottom left)} the coalescence time method and
            \textbf{(bottom right)} the resistance distance method.
        \label{sfig:expansion_results}
    }
\end{figure}

\begin{figure}
\centering
    \includegraphics[width=0.65\textwidth]{poplars/posterior_dists_g_populus}
    \includegraphics[width=0.65\textwidth]{poplars/grid_populus}
    \caption{
        Posterior distributions of parameters for \textit{Populus} data,
        using coalescence time inference with the model shown in Figure~\ref{fig:poplar_results}.
        \textbf{(top)} Posterior medians (dark line),
        interquartile range (box), and middle 95\% (whiskers),
        for gene flow rates.
        \textbf{(bottom)} Posterior medians, along with edge labels that match the upper figure.
        \label{sfig:poplar_g}
    }
\end{figure}
